Relationships in the extremes and their influence on competition and coexistence

### 2022-04-08 paper/Paper.Rnw
minor edits to discussion

#place R code here for loading in necessary variables from Results
#ordinarily this would be a readRDS command, but I'm just defining a variable here for now
#allregres<-matrix(rnorm(10),2,5)
fig2 <- readRDS(file="../results_numeric/fig2data.RDS")
M <- readRDS(file="../results_numeric/M.RDS")
se3 <- readRDS(file="../results_numeric/fig3maxse.RDS")
se4 <- readRDS(file="../results_numeric/fig4maxse.RDS")
@

@@

%future work: study ATAs in environmental variables and how theyre changing - tie in with specialist
Our study highlights the importance of ATAs for ecological phenomena 
and specifically for coexistence and the maintenance of biodiversity. 
The occurence of ATAs among environmental variables themselves, 
their contribution to coexistence, and their changes through space and time
can and should be explored further. 
The coupling of the magnitude of ATA contribution and environmental variability, 
as shown in our results and figures (Fig. \ref{fig:fig3} - \ref{fig:fig6}), 
suggests that ATAs and their effects
may become more important to consider for coexistence and possibly other 
ecological phenomena as environmental conditions evidently
become more variable.
As described in section \ref{section:introduction}, changes in the ATAs between environmental
variables may occur as the climate becomes more variable and extreme events become more 
expansive and extensive, causing relationships between separated events to become tighter at one end or another and thus asymmetric. Additionally, 
if ATA specialist exist, their persistance may
become at stake as the mechanism driving their coexistence changes with the climate.
Thus it is of interest to further study how tail associations of environmental variables are changing and how this might 
alter competition, coexistence, and the maintenance of biodiversity generally.
Our investigation brings to attention the consequences of such
possible \emph{relationships} in the extremes; however, 
simpler changes in [the marginal/ univariate distributions of] extreme events themselves such as
intensity and frequency and their influence on coexistence also merit future work. 
Possible changes in tail association
patterns between environmental variables in nature can
influence the coexistence of species through the storage effect and not other mechanisms, but
extreme and changing climate is probable to affect other mechanims in other ways.
%In the Discussion, I intend to talk about how possible changes in tail association 
%patterns of env vars in nature may influence coexistence of species. It won't influence the 
%other mechanisms, though changes in the extreme values of environmental variables, known
%to be occurring, might. 
%Should also mention this second point in the Discussion as something that merits further research.
%though might be worth mentioning in the Discussion that events are getting more extreme. We have studied how *relationships*
%in the extremes influence coexistence, but probably worth studying (or cite papers if it's been done) how changes
%in the commonness and extremeness of extreme events themselves influence coexistence, and for that you'd have
%to also look at the other mechanisms. May still need a sup mat laying out all the math.

@@

%\underline{E and vital rates}
We note that \emph{E}, in our investigation and in other investigations within the MCT 
framework, is not directly a measure of the environment, but is a measure of response to the 
environment, measured by a population's vital rates. Environment and environmental response may be related linearly
(like lottery model?); though real environmental variables are probable to 
actually be related to vital rates in a non-linear and often threshold-like way. 
Threshold relationships can produce tail association, as we showed elsewhere (explain, ref). 
Changes in the means of true environmental variables may change 
where those distributions of environmental variables are compared to thresholds, 
which may change the nature of tail associations between the effects of 
those environmental variables. Thus, not only do the possible asymmteric association between 
environmental variables merit future study, but the relationships between populations'
vital rates should also be investigated for species coexistence and the maintenacne of biodiversity.
%Discussion content: We should (in future) study tail associations between environmental variables, and how these might be changing. However, more than that, the variables E in the lottery model are not actual environmental variables but rather are the effects of environmental variables on the species of the model.

%\underline{the new effects are not "weak noise" effects}
%if sigma really small, ATA would not matter?
%go back to emails
%Maybe have a paragraph about how Chesson's weak noise and other assumptions make it impossible to detect this new effect
%via that framework. It's only the less-assumption-reliant simulation approach of Ellner et al that makes it possible to 
%study this new term in the decomp. Need to study a bit to substantiate this properly, or get Steve Ellner help.
We parse the storage effect, providing a tertiary decomposition to the simulation-based 
approach of MCT \citep{Ellner_2016, Ellner_2019}, 
a new approach which bypasses the mathematical assumptions and obstructions of 
standard MCT. Classical analytic derivations of MCT using second order Taylor expansion and 
small-variance approximations completely miss out
on any additions to coexistence beyond simple correlations and does not include our new term, $\Delta^{[EC]}$ or $\Delta^{(E \shp C)}$ \citep{Chesson_1994, Chesson_2000}. 
In fact, storage effects in analytic MCT is computationally equal to $\Delta^{[E \nat C]}$, 
the contributions of correlation \emph{per se}. 
In newer, simulation based MCT, storage effect is 
the sum of the two new terms, $\Delta^{(EC)}=\Delta^{[E \nat C]}+\Delta^{[EC]}$, 
obscuring the effects of ATAs. However, because standard MCT's assumption make it impossible to 
detect interactions outide of correlation \emph{per se}, simulation-based MCT is required to
detect ATA contributions, and our new extension to it cleary separates and quantifies the 
relative contribution and importance of ATAs and pure correlation to the storage effect
and subsequently species coexistence and the maintenance of biodiversity.

@@

The chances of ATAs being important for the coexistence of species increases as more variables are included in the system, 
either through the addition of more species or more drivers (beyond just E and C), or both. 
When multiple species are included in the system, the interaction between different species' E's and C's introduces more relationships 
to consider that may or may not be asymmetrically related and could increase the relative importance of ATA contribution to the species' coexistence. 
Likewise, coexistence between species can also depend on more than two drivers \citep{Ellner_2019}. 
The storage effect decomposition presented here is generalizable to coexistence decompositions of multiple variables.
Following the set up and notation of \cite{Ellner_2019}'s \emph{covariance decomposition}, 
ATA contributions and correlation \emph{per se} can be decomposed from the \emph{associations} between two variables
among many from the appropriate term with parenthesized superscripts with the $\shp$ symbol denoting and separating 
variables that are varying independently from the variables that are covarying whose relationship we are interested in decomposing. 
For example, the term $\Delta^{(XY \shp Z)}$ could be decomposed into $\Delta^{[XY \shp Z]} + \Delta^{[X \nat Y \shp Z]}$. 
As more terms and species are introduced into a system, more relationships between variables must be evaluated,
more of those relationships may be asymmetrically tail associated, and ATAs may contribute more to the maintenance of 
biodiversity in that system. 
%make three different biv dis that are standard normal copula and then pair with a shuffles third
%use to get rbar and subtract from rbar of XY#Z,

@@

%Also need to talk about how this stuff will apply to data.
As an introductory investigation of ATAs within the framework of MCT, 
we used simple and extreme examples, and future work should assess more complex relationships, systems, and models. 
In our theoretical exploration, we implemented a simple lottery model and assessed 
the influence of extremely asymmetric associated noise, i.e., the noise was perfectly correlated
in exactly one half of the distribution. However, more complex models (such as structured demographic models
or integral projection models should be explored )and varying degrees and qualities of ATA, (copula parameters, 
difference between tails, rate of relatedness(slope of expansion??) should be implemented and explored.
Moreover, because our empirical application makes use of a simple and controlled lab system, future studies
should be done to assess the extent of how much ATA are important for complex systems and field systems.
Because modeling the dynamics of long-lived populations can be challenging, a good field system to start
exploring the ideas presented here is a system of annual plants. 
For example, the ATAs between rainfall timing on the coexistence of grasses and forbs 
in California grasslands \citep{Hallett_2019} could be a fruitful 
further investigation of biodiversity maintenance in California grasslands and the effects of extreme 
rainfall fluctuation given the highly variable climate of the area.
Moreover, our approach may provide more insight when expanded to consider fluctuations across space,
in regards to questions about population synchrony, Moran effects, and the suggestion of possible ATA specialists.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig5.pdf}
\caption{The contribution of ATAs to GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
other components of the MCT decomposition. All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
For each panel, 
two of the three parameters defining the temperature oscillation (the amplitude, $a$; period, $P$; 
and mean, $\theta_0$; see Methods) were fixed at the values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, $\theta_0=18^\circ$C), and 
GWR and components of the MCT decomposition
were then plotted against the third parameter. The original values of the varying parameters 
are indicated by vertical dashed magenta lines. ATA contributions 
had magnitude comparable to other components, 
indicating that ATAs can contribute meaningfully
to diatom coexistence in this system. In some instances, the GWR line (orange) falls 
between the ATA line (red)
and the horizontal axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of \emph{F. crotonensis}. 
Standard errors of plotted quantities were always less then X, so are not shown.
%***DAN: Don't forget to fill in the standard error stuff, please, Jasmin.
For the parameters we considered, GWR of the other species, \emph{C. pseudostelligera}, 
was always positive both including and
excluding the effects of ATAs, so GWR for \emph{F. crotonensis}, decomposed here, is a 
sufficient coexistence metric.}  \label{fig:fig5}
\end{figure}

%***DAN: Jasmin, some changes to this figure are needed. Please make axis labels and numbers bigger. Please put 
%axis labels outside the axes instead of inside. Please make panel labels bigger. Please make coutour labels
%slightly bigger.
%***DAN: Do we want to do anything with standard error for this fig?
\begin{figure}
\includegraphics[width=.38\textwidth]{../results_figs/fig6.pdf}
\caption{The contribution of ATAs to GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
the GWR itself. The ratio $\Delta_i^{[EC]}/\text{GWR}$ is shown, larger magnitudes
indicating greater importance of ATAs and values greater than $1$ corresponding to either
ATA rescue or ATA exclusion (Methods). For this system and the plotted parameters, 
$\Delta_i^{[EC]}$ was always positive, so cyan indicates exclusion of \emph{F. crotonensis} and
magenta indicates coexistence of the two diatom species. Values greater than 1 or less than -1
were remapped to 1 and -1, respectively, for readability.
For the parameters we considered, GWR of the other species, \emph{C. pseudostelligera}, was always positive both including and
excluding ATAs, so GWR for \emph{F. crotonensis} is a sufficient coexistence metric.} \label{fig:fig6}
\end{figure}

>>>>

#place R code here for loading in necessary variables from Results
#ordinarily this would be a readRDS command, but I'm just defining a variable here for now
#allregres<-matrix(rnorm(10),2,5)
res_loc <- "../results_numeric/"
fig2 <- readRDS(file="../results_numeric/fig2data.RDS")
M <- readRDS(file="../results_numeric/M.RDS")
se3 <- readRDS(file="../results_numeric/fig3maxse.RDS")
se4 <- readRDS(file="../results_numeric/fig4maxse.RDS")
se5 <- readRDS(file=paste0(res_loc, "fig5maxse.RDS"))
se6 <- readRDS(file=paste0(res_loc, "fig6maxse.RDS"))
@

@@

%future work: study ATAs in environmental variables and how theyre changing - tie in with specialist
Our study highlights the importance of ATAs for ecological phenomena 
and specifically for coexistence and the maintenance of biodiversity. 
The occurence of ATAs among environmental variables themselves, 
their contribution to coexistence, and their changes through space and time
can and should be explored further. 
The coupling of ATA contribution magnitude with environmental variability, 
as shown in our results and figures (Fig. \ref{fig:fig3} - \ref{fig:fig6}), 
suggests that ATAs and their effects
may become more important to consider for coexistence and possibly other 
ecological phenomena as environmental conditions may evidently
become more variable (probably need refs).
As described in section \ref{section:introduction}, ATAs between environmental
variables may shift as the climate becomes more variable and extreme events become more 
expansive and extensive, causing associations of distinct temporal events to become 
more related in one extreme or tail, thus asymmetric. Additionally, 
if ATA specialist exist, their persistance may
become at stake as the mechanism driving their coexistence changes with the climate.
Thus it is of interest to further study how tail associations of environmental variables are changing and how this might 
alter competition, coexistence, and the maintenance of biodiversity generally.
Our investigation brings to attention the consequences of such
possible [bivariate] \emph{relationships} in the extremes; however, 
simpler changes in [the marginal/ univariate distributions of] extreme events themselves such as
intensity and frequency and their influence on coexistence also merit future work. 
Possible changes in tail association
patterns between environmental variables in nature can
influence the coexistence of species through the storage effect and not other mechanisms, but
extreme and changing climate is probable to affect other mechanims in other ways (refs?).
%In the Discussion, I intend to talk about how possible changes in tail association 
%patterns of env vars in nature may influence coexistence of species. It won't influence the 
%other mechanisms, though changes in the extreme values of environmental variables, known
%to be occurring, might. 
%Should also mention this second point in the Discussion as something that merits further research.
%though might be worth mentioning in the Discussion that events are getting more extreme. We have studied how *relationships*
%in the extremes influence coexistence, but probably worth studying (or cite papers if it's been done) how changes
%in the commonness and extremeness of extreme events themselves influence coexistence, and for that you'd have
%to also look at the other mechanisms. May still need a sup mat laying out all the math.

@@

%\underline{E and vital rates}
We note that \emph{E}, in our investigation and in other investigations within the MCT 
framework, is not directly a measure of the environment, but is a measure of response to the 
environment, via a population's vital rates. Environment and environmental response may be related linearly
(like lottery model?); though real environmental variables are probable to 
actually be related to vital rates in a non-linear and often threshold-like way. 
Threshold relationships can produce tail association, as we showed elsewhere (explain, ref). 
Changes in the means of true environmental variables may change 
where those distributions of environmental variables are compared to thresholds, 
which may change the nature of tail associations between the effects of 
those environmental variables. Thus, not only do the possible asymmteric association between 
environmental variables merit future study, but the relationships between populations'
vital rates should also be investigated for species coexistence and the maintenance of biodiversity.
%Discussion content: We should (in future) study tail associations between environmental variables, and how these might be changing. However, more than that, the variables E in the lottery model are not actual environmental variables but rather are the effects of environmental variables on the species of the model.

%\underline{the new effects are not "weak noise" effects}
%if sigma really small, ATA would not matter?
%go back to emails
%Maybe have a paragraph about how Chesson's weak noise and other assumptions make it impossible to detect this new effect
%via that framework. It's only the less-assumption-reliant simulation approach of Ellner et al that makes it possible to 
%study this new term in the decomp. Need to study a bit to substantiate this properly, or get Steve Ellner help.
We parse the storage effect, providing a tertiary decomposition to the simulation-based 
approach of MCT \citep{Ellner_2016, Ellner_2019}, 
a new approach which bypasses the mathematical assumptions and obstructions of 
standard MCT. Classical analytic derivations of MCT using second order Taylor expansion and 
small-variance approximations completely miss out
on any additions to coexistence beyond simple correlations and does not include our new ATA term, $\Delta^{[EC]}$, or variance \emph{per se}, $\Delta^{(E \shp C)}$ \citep{Chesson_1994, Chesson_2000}. 
In fact, storage effects in analytic MCT is computationally equal to $\Delta^{[E \nat C]}$, 
the contributions of correlation \emph{per se}. 
In newer, simulation based MCT, storage effect is 
the sum of the two new terms, $\Delta^{(EC)}=\Delta^{[E \nat C]}+\Delta^{[EC]}$, 
obscuring the effects of ATAs. Because standard MCT's assumption make it impossible to 
detect interactions outide of correlation \emph{per se}, simulation-based MCT is required to
detect ATA contributions, and our new extension to it clarifies and cleary separates and quantifies the 
relative importance of both ATAs and pure correlation in contribution to the storage effect
and subsequently species coexistence and the maintenance of biodiversity.

@@

The chances of ATAs being important for the coexistence of species increases as more variables are included in the system, 
either through the addition of more species or more drivers (beyond just E and C), or both. 
When multiple species are included in the system, the interaction between different species' E's and C's introduces more relationships 
to consider that may or may not be asymmetrically related and could increase the relative importance of ATA contribution to the species' coexistence. 
Likewise, coexistence between species can also depend on more than two drivers \citep{Ellner_2019}. 
The storage effect decomposition presented here is generalizable to coexistence decompositions of multiple variables.
Following the set up and notation of \cite{Ellner_2019}'s \emph{covariance decomposition}, 
ATA contributions and correlation \emph{per se} can be decomposed from the \emph{associations} between two variables
among many from the appropriate term with parenthesized superscripts containing the $\shp$ symbol which denotes and separates 
variables that are varying independently from the variables that are covarying whose relationship we are interested in decomposing. 
For example, the term $\Delta^{(XY \shp Z)}$ could be decomposed into $\Delta^{[XY \shp Z]} + \Delta^{[X \nat Y \shp Z]}$. 
As more terms and species are introduced into a system, more relationships between variables must be evaluated,
more of those relationships may be asymmetrically tail associated, and ATAs may contribute more to the maintenance of 
biodiversity in that system. 
%make three different biv dis that are standard normal copula and then pair with a shuffles third
%use to get rbar and subtract from rbar of XY#Z,

@@

%Also need to talk about how this stuff will apply to data.
As an introductory investigation of ATAs within the framework of MCT, 
we used simple and extreme examples, and future work should assess more complex relationships, systems, and models. 
In our theoretical exploration, we implemented a simple lottery model and assessed 
the influence of extremely asymmetric associated noise, i.e., the noise was perfectly correlated
in exactly one half of the distribution. However, more complex models (such as structured demographic models
or integral projection models should be explored )and varying degrees and qualities of ATA, (copula parameters, 
difference between tails, rate of relatedness(slope of expansion??) should be implemented and explored.
Moreover, because our empirical application makes use of a simple and controlled lab system, future studies
should be done to assess the extent of how much ATAs are important for complex systems and field systems which are 
often subject to factors beyond our control.
Because modeling the dynamics of long-lived populations can be challenging, a good field system to start
exploring the ideas presented here is a system of annual plants. 
For example, the ATAs effcts of rainfall timing on the coexistence of grasses and forbs 
in California grasslands \citep{Hallett_2019} could be a fruitful 
further investigation of biodiversity maintenance in California grasslands and the effects of extreme 
rainfall fluctuation given the highly variable climate of the area.
Moreover, our approach may provide more insight when expanded to consider fluctuations across space,
in regards to questions about population synchrony, Moran effects, and the suggestion of possible ATA specialists.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig5.pdf}
\caption{The contribution of ATAs to GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
other components of the MCT decomposition. All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
For each panel, 
two of the three parameters defining the temperature oscillation (the amplitude, $a$; period, $P$; 
and mean, $\theta_0$; see Methods) were fixed at the values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, $\theta_0=18^\circ$C), and 
GWR and components of the MCT decomposition
were then plotted against the third parameter. The original values of the varying parameters 
are indicated by vertical dashed magenta lines. ATA contributions 
had magnitude comparable to other components, 
indicating that ATAs can contribute meaningfully
to diatom coexistence in this system. In some instances, the GWR line (orange) falls 
between the ATA line (red)
and the horizontal axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of \emph{F. crotonensis}. 
Standard errors of plotted quantities were always less then \Sexpr{round(se5,5)}, so are not shown.
For the parameters we considered, GWR of the other species, \emph{C. pseudostelligera}, 
was always positive both including and
excluding the effects of ATAs, so GWR for \emph{F. crotonensis}, decomposed here, is a 
sufficient coexistence metric.}  \label{fig:fig5}
\end{figure}

%***DAN: Jasmin, some changes to this figure are needed. Please make axis labels and numbers bigger. Please put 
%axis labels outside the axes instead of inside. Please make panel labels bigger. Please make coutour labels
%slightly bigger.
%***DAN: Do we want to do anything with standard error for this fig?
\begin{figure}
\includegraphics[width=.38\textwidth]{../results_figs/fig6.pdf}
\caption{The contribution of ATAs to GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
the GWR itself. The ratio $\Delta_i^{[EC]}/\text{GWR}$ is shown, larger magnitudes
indicating greater importance of ATAs and values greater than $1$ corresponding to either
ATA rescue or ATA exclusion (Methods). For this system and the plotted parameters, 
$\Delta_i^{[EC]}$ was always positive, so cyan indicates exclusion of \emph{F. crotonensis} and
magenta indicates coexistence of the two diatom species. Values greater than 1 or less than -1
were remapped to 1 and -1, respectively, for readability.
For the parameters we considered, GWR of the other species, \emph{C. pseudostelligera}, was always positive both including and
excluding ATAs, so GWR for \emph{F. crotonensis} is a sufficient coexistence metric.
Standard errors of plotted quantities were always less then \Sexpr{round(se6,5)}, so are not shown.} \label{fig:fig6}
\end{figure}
